Evolutionary rescue and reintroduction of resistant frogs allows recovery in the presence of a lethal fungal disease

### 2023-12-02 doc/manuscript/translocation.qmd
Changes to text in response to reviewer comments

---
bibliography: translocation.bib
csl: pnas.csl
# format: docx
format :
  # latex:
  pdf:
    cite-method: natbib
    include-in-header:
      text: |
        \usepackage{blkarray}
editor: visual
editor_options:
  markdown:
    wrap: sentence
---

@@

One consequence of human alteration of the biosphere is an increase in emerging infectious diseases [@jones2008; @fisher2012].
Such diseases pose a severe threat to wildlife populations [@daszak2000], and have caused dramatic declines and extinctions in a wide range of taxa, including echinoderms, mammals, birds, and amphibians [@hewson2014; @samuel2015; @scheele2019; @cunningham2021].
Amid this global loss of biodiversity, some populations decimated by disease have subsequently recovered to varying extents [@scheele2017; @voyles2018; @knapp2016].
Although we know relatively little about the mechanisms underlying recovery [@brannelly2021; @russell2020], host evolution is one possibility [@carlson2014; @searle2020].
This process is known from a diversity of wildlife populations [@savage2016; @epstein2016; @gignoux-wolfsohn2021; @holland2022], and often results in increased host resistance and/or tolerance [see @medzhitov2012 for definitions].
However, linking these evolutionary responses to population recoveries is difficult and few clear examples exist.
In addition to promoting recovery of disease-impacted populations, the resistant/tolerant individuals in such populations could potentially be used to reestablish populations that were extirpated following disease emergence [@mendelson2019; @brannelly2021].
To date, the use of resistant/tolerant individuals for species recovery is rare [but see @joseph2018], and studies that evaluate its potential utility are critically needed.

@@

Despite the catastrophic impact of Bd on MYL frogs, wherein most Bd-naive populations are extirpated following Bd arrival [@vredenburg2010], some populations have persisted after epizootics [@briggs2010] and are now recovering [@knapp2016].
Frogs in these recovering populations show reduced susceptibility to Bd infection, consistent with selection for more resistant/tolerant genotypes [@knapp2016].
Furthermore, exploratory efforts to reintroduce frogs collected from recovering populations to sites from which frogs were previously extirpated indicate that population establishment is possible despite ongoing Bd infection [@joseph2018].

In the current study, we combine results from genomic analyses, a 15-year landscape-scale frog reintroduction effort, and a detailed population model to provide an unprecedented example of how evolution in response to disease allows the successful reestablishment of extirpated MYL frog populations.
Specifically, we used exome capture methods to show that MYL frogs have undergone substantial evolutionary change associated with Bd exposure, changes that might provide a genetic basis for the increased resistance/tolerance to Bd infection observed in naturally recovering populations [@knapp2016].
We then use the reintroduction of resistant/tolerant frogs to 24 sites and results from long-term capture-mark-recapture (CMR) surveys to describe the dynamics and drivers of population establishment.
These data are also used to parameterize a population model that extends our inferences of population establishment well beyond the temporal extent of the CMR data.
Collectively, these results provide one of the first examples of how evolutionary rescue and the reintroduction of resistant/tolerant animals can allow the large-scale recovery of disease-impacted species.

@@

@@

Reintroduction of resistant/tolerant MYL frogs was remarkably successful in reestablishing viable populations in the presence of Bd.
Of the 12 translocated populations, approximately 80% showed evidence of both successful reproduction and recruitment of new adults.
Bd loads measured before and after translocation were similar and generally remained well below the level at which frog mortality occurs.
Annual survival for 12 of the 24 translocated cohorts exceeded 50%, and \> 70% of translocated cohorts had survival above this 50% level when the earliest translocations are excluded (i.e., translocations conducted when methods were still being refined).
The fact that the relatively low Bd loads and correspondingly high frog survival in the presence of Bd was maintained between the natal/donor sites and the recipient sites indicates that these characteristics of naturally-recovering populations were not solely an effect of the natal/donor site or site \* frog interactions, but instead were due at least in part to resistance/tolerance inherent in the frogs.
In addition, the relatively high survival of translocated frogs was maintained in their progeny, as expected if resistance/tolerance has a genetic basis.
Results from the population viability model were also encouraging.
In particular, translocated populations with \> 50% survival in the first year post-translocation were predicted to have a low probability of extinction over 50 years (probability of extinction = 0 - 0.32).
In light of the generally low success rate of amphibian reintroduction efforts [@dodd2005], our success in reestablishing MYL frog populations via translocation of resistant/tolerant individuals is striking, and even more so given that this success occurred despite ongoing infection of frogs with Bd.

Despite the resistance/tolerance of adult MYL frogs in recovering populations, individual and population-level impacts of Bd are still evident.
In an earlier study of 2 of our 12 translocated populations [@joseph2018], Bd infection and load had detectable effects on the survival of adults and may have influenced population establishment (sites referred to as "Alpine" and "Subalpine" in @joseph2018 are identified as "70550" and "70505" in the current study).
Applying similar analyses to all 12 of our translocated populations would likely provide a broader perspective of the ongoing effect of Bd.
In addition to these important but relatively subtle effects of Bd on adults, the impacts on younger life stages are more apparent.
MYL frogs immediately following metamorphosis ("metamorphs") are highly susceptible to Bd infection [@ellison2018] and as a result experience high mortality.
This high susceptibility of metamorphs is documented in numerous species of anurans, and may result from the poorly developed immune system characteristic of this life stage [@humphries2022].
In naturally recovering and translocated MYL frog populations, we suggest that the high mortality of metamorphs is an important limitation on subsequent recruitment of new adults.
Therefore, although adult MYL frogs appear relatively resistant/tolerant, Bd infection continues to have important limiting effects on recovering populations [see also @hollanders2022].

@@

We conducted `r transno_min`--`r transno_max` translocations per site (@fig-yosemap, @fig-translocation-survival) and each translocated cohort included `r transfrogs_min` to `r transfrogs_max` frogs (median = `r transfrogs_median`).
In preparation for each translocation, adult frogs were collected from the donor population and measured, weighed, swabbed, and PIT tagged.
Frogs were transported to the recipient site either on foot or via helicopter.
Following release, translocated populations were visited approximately once per month during the summer active season (median number of visits per summer = `r round(surveys_yr_median)`; range = `r surveys_yr_min`--`r surveys_yr_max`) and assessed using diurnal CMR surveys and VES.
CMR surveys allowed estimation of adult survival, adult population size, and recruitment of new adults, and VES provided estimates of tadpole and juvenile abundance.
During 2006-2012, CMR surveys were conducted on a single day (primary period) per site visit, during which all habitats were searched repeatedly for adult frogs.
Frogs were captured using handheld nets, identified via their PIT tag (or tagged if they were untagged), measured, weighed, swabbed, and released at the capture location.
During 2013-2022, we generally used a robust design in which all habitats were searched during several consecutive days (median number of secondary periods per primary period = `r round(robust_year_median)`; range = `r robust_year_min`--`r robust_year_max`), and frogs were processed as described above.
However, when the number of frogs detected on the first survey day was zero or near zero, we typically conducted only a single-day CMR survey.
When using a robust design, within a primary period, frogs that were captured during more than one secondary period were measured, weighed, and swabbed during the first capture, and during subsequent captures were only identified and released.

@@

For each of the translocated populations, we used the presence of tadpoles and/or juveniles from VES and counts of new recruits (i.e, untagged adults) in CMR surveys to provide two measures of successful reproduction.
To calculate the proportion of years in which tadpoles/juveniles were present at a site, we excluded surveys conducted in the year of the initial translocation to that site.
This exclusion accounted for the fact that all translocations were conducted after the breeding period and reproduction would therefore not have occurred until the following year.
Similarly, to calculcate the proportion of years in which new recruits were present at a site, we excluded surveys conducted during the 3 years following the initial translocation.
This accounted for the multi-year tadpole and juvenile stages in MYL frogs ([@tbl-param_values]).

@@

![Evidence for selection on individual variants in recovering MYL frog populations at the landscape scale. For each of the 9 naive and recovering MYL frog populations (indicated by numbered points), adjacent pie charts show allele frequencies for the 11 outlier SNPs from 7 distinct genes: (A) LOC108802036, (B) TCF19, VARS, MAP1S, TYRP1, and SRFBP1, and (C) RIN3. Charts are superimposed on a map of the Sierra Nevada study area, with Yosemite, Kings Canyon, and Sequoia National Parks (from north to south) shown in dark gray, and the range boundary for MYL frogs shown in light gray. The inset map locates the study area and range boundary in California.](figures/allele_maps.png){#fig-allelefrequencies width="450"}

@@

@@

We compared frog genomes sampled in naive versus recovering populations.
Bd-exposure histories of these populations are based on a decade or more of data from visual encounter surveys and Bd surveillance using skin swabbing [e.g., @knapp2016].
Comparing populations with different infection histories allowed larger sample sizes and replication across the landscape.
The alternative approach of comparing samples from the same populations before and after Bd exposure isn't feasible in this system because Bd arrived in most MYL frog populations decades ago and population persistence/recovery is rare and unpredictable.
As a result, samples from recovering populations collected from before and after Bd exposure are not available and are unlikely to be available in the future.

#### Swab Methods

@@

#### Mark-recapture model structure

We estimated survival and recruitment for each site using open population mark recapture models based on @joseph2018.
For each individual $i=1, ..., M$ on each survey $j=1, ..., n_j$: $o_{i, j} = 1$ if the individual was not detected, and $o_{i, j}=2$ if the individual was detected. 
Capture histories of $M$ individuals are modeled, though only $N_s$ individuals were captured.
This parameter expanded data augmentation allows us to capture the possibility that undetected individuals may have recruited into the adult population [@royle2012parameter]. 
Here, $M$ was chosen to be three times the number of observed individuals ($3N_s$) to be considerably greater than our prior guess of $N_s$.

We denote the true state of individual $i$ as $u_{i, t}$ for primary period $t = 1,..., n_t$. 
The four states that we consider are: $u_{i, t} = 1$ for individuals that have not recruited, $u_{i, t} =2$ for live adults, and $u_{i, t} = 3$ for dead adults. 
Each survey $j=1, ..., n_j$ occurs in one of the $n_t$ primary periods, and we denote the primary period in which survey $j$ takes place as $t_j$. 
Each primary period $t$ occurs within one year, but within a year there are multiple primary periods.
We set the year containing the first primary period to $y_{t = 1} = 1$, and generally $y_t$ represents the year containing primary period $t$. 
Years increment by one until the final year of the mark recapture efforts, which we denote $n_y$: $y \in \{1, 2, ..., n_y\}$.
We assume that within a primary period, the state of each individual does not change (i.e., individuals do not recruit into the adult population, gain or lose Bd infection, or die). 
This assumption is justified by the short time intervals between surveys within primary periods.

Live individuals are detected with probability $p_j$, which is modeled as:

$$p_j = \text{logit}^{-1}(X_j^{(p)} \beta^{(p)}),$$

where $X_j^{(p)}$ is a known row vector and $\beta^{(p)}$ an unknown parameter vector. 
Not recruited and dead individuals are never captured. 
We bundle these assumptions about the observation probabilities for survey $j$ into an emission matrix $\Omega_j$:

$$
\Omega_j =
\begin{blockarray}{ccc}
  \text{Not detected} & \text{Detected} \\
\begin{block}{(cc)c}
  1 & 0 & \text{Not recruited} \\
  1 - p_j & p_j & \text{Alive} \\
  1 & 0 & \text{Dead} \\
\end{block}
\end{blockarray}
$$

The state transition matrix $\Psi_{t, i}$ contains the probabilities of individual $i$ transitioning from state $u_{i, t}$ (rows) to $u_{i, t+1}$ (columns) between primary period $t$ and $t+1$. 
For unintroduced individuals which naturally recruit, this matrix is given by:

$$
\Psi_{t, i} =
\begin{blockarray}{cccc}
  \text{Not recruited} & \text{Alive} & \text{Dead} \\
\begin{block}{(ccc)c}
  1 - \lambda_t & \lambda_t & 0 & \text{Not recruited} \\
  0 & \phi_t & 1 - \phi_t & \text{Alive} \\
  0 & 1 & 1 & \text{Dead} \\
\end{block}
\end{blockarray},
$$
where $\lambda_t$ is the probability of recruiting in time $t$ and $\phi_t$ is the probability of survival in time $t$.

For introduced individuals which have deterministic recruitment (they recruit when introduced), the state transition matrix is given by:

$$
\Psi_{t, i} =
\begin{blockarray}{cccc}
  \text{Not recruited} & \text{Alive} & \text{Dead} \\
\begin{block}{(ccc)c}
  1 - I_{t, i} & I_{t, i} & 0 & \text{Not recruited} \\
  0 & \phi_t & 1 - \phi_t & \text{Alive} \\
  0 & 1 & 1 & \text{Dead} \\
\end{block}
\end{blockarray},
$$
where $I_{t, i}$ is a known indicator function for whether individual $i$ was introduced in primary period $t$.

We allow recruitment probabilities to vary in time via random effects, such that:

$$\lambda_t = \text{logit}^{-1}(\alpha^{(\lambda)} + \epsilon^{(\lambda)}_t),$$

where $\alpha^{(\lambda)}$ is an intercept parameter and $\epsilon^{(\lambda)}_t$ is an adjustment for time $t$.

Survival probabilities also vary in time, and as a function of known covariates:

$$\phi_t = \text{logit}^{-1}(X^{(\phi)}_t \beta^{(\phi)} + \epsilon^{(\phi)}_t),$$

where $X^{(\phi)}_t$ is a row vector of known covariates, $\beta^{(\phi)}$ is a column vector of unknown coefficients, and $\epsilon^{(\phi)}_t$ is an adjustment for time $t$.

To complete the specification of the Bayesian model, we need to specify priors for all unknown parameters. 
The recruitment parameter priors were specified as follows:

$$\alpha^{(\lambda)} \sim N(0,1),$$
$$\sigma^{(\lambda)} \sim N_+(0,1),$$
$$\epsilon^{(\lambda)}_t \sim N(0, \sigma^{(\lambda)}),$$
for periods $t=1, ..., T$. 
Here $N$ represents the normal distribution and $N_+$ the half normal distribution with positive support.

Survival parameter priors were specified similarly as:

$$\beta^{(\phi)} \sim N(0, 1),$$
$$\sigma^{(\phi)} \sim N_+(0, 1),$$
$$\epsilon^{(\phi)}_t \sim N(0, \sigma^{(\phi)}),$$
for time $t=1, ..., T$.

The detection model coefficient vector also received a standard normal prior $\beta^{(p)} \sim N(0, 1)$.

We computed the likelihood of each individual capture history using the forward algorithm, and we estimated the latent states using the forward-backward algorithm (@zucchini2009hidden, @joseph2018). 
All of the code to specify and fit the model in Stan is available in the open source mrmr package (@joseph2019).

The joint distribution of the resulting model can be written as follows:

$$
\begin{multline*}
[\alpha^{(\lambda)}, \sigma^{(\lambda)}, \epsilon^{(\lambda)}_{1:T}, \beta^{(\phi)}, \sigma^{(\phi)}, \epsilon^{(\phi)}_{1:T}, \beta^{(p)} \mid \pmb Y] \propto \prod_{i=1}^M [Y_i \mid \alpha^{(\lambda)}, \epsilon^{(\lambda)}_{1:T}, \beta^{(\phi)}, \epsilon^{(\phi)}_{1:T}, \beta^{(p)}] \times \\
\prod_{t=1}^T [\epsilon_t^{(\lambda)} \mid \sigma^{(\lambda)}] [\epsilon_t^{(\phi])} \mid \sigma^{(\phi)}] [\sigma^{(\lambda)}] [\sigma^{(\phi)}] [\alpha^{(\lambda)}] [\beta^{(\phi)}] [\beta^{(p)}],
\end{multline*}
$$

where $\pmb Y$ is an $M \times T$ detection matrix, and $Y_i$ the capture history of individual $i$.

@@

![Map showing the locations of translocated and donor MYL frog populations in Yosemite National Park (park boundary indicated by black polygon). Symbol shapes indicate the donor population used for each translocation site. To obscure the exact locations of populations, random noise was added to all point coordinates. Inset map shows the location of Yosemite within California. In both maps, elevation is indicated by the colored hillshade layer (dark green = lowest elevation, white = highest elevation).](figures/map_translocation_points.png){#fig-yosemap width="250"}

@@

![Evidence that frogs translocated from recovering populations can maintain the benefits of resistance/tolerance in non-natal habitats. For frogs translocated to each of the 12 recipient sites, Bd loads for the period immediately prior to translocation versus during the 1-year period after translocation. Bd loads are expressed as the number of ITS1 copies per skin swab, as estimated by qPCR of the Bd ITS1 region. Box plots show medians, first and third quartiles, largest and smallest values within 1.5x interquartile range, and values outside the 1.5x interquartile range. Loads indicative of severe disease are \> 5.8 ITS copies (on a log~10~ scale).](figures/bdload_beforeafter.png){#fig-bdload-beforeafter width="500"}

@@

| **Parameter**                                                             | **Value**            | **Source**                                                         |
|:-----------------------------|:---------------|:--------------------------|
| $\sigma_{L_1}$, Yearly survival probability of year 1 tadpoles            | 0.7                  | Estimated from field data, observations, natural history knowledge |
| $\sigma_{L_2}$, Yearly survival probability of year 2 tadpoles            | 0.7                  | Estimated from field data, observations, natural history knowledge |
| $\sigma_{L_3}$, Yearly survival probability of year 3 tadpoles            | 0.7                  | Estimated from field data, observations, natural history knowledge |
| $\sigma_{J_1}$, Yearly survival probability of year 1 juveniles           | 0.25                 | Estimated from field data, observations, natural history knowledge |
| $\sigma_{J_2}$, Yearly survival probability of year 2 juveniles           | 0.5                  | Estimated from field data, observations, natural history knowledge |
| $\sigma_{A_R}$, Yearly survival probability of naturally recruited adults | Varies by population | Estimated from mark-recapture studies                              |
| $\sigma_{A_T}$, Yearly survival probability of translocated adults        | Varies by population | Estimated from mark-recapture studies                              |
| $p_{L_1}$, Probability of a year 1 tadpoles remaining as a tadpoles       | 1                    | Estimated from field data, observations, natural history knowledge |
| $p_{L_2}$, Probability of a year 2 tadpoles remaining as a tadpoles       | 0.25                 | Estimated from field data, observations, natural history knowledge |
| $p_{J_1}$, Probability of a year 1 juvenile remaining as a juvenile       | 0.25                 | Estimated from field data, observations, natural history knowledge |
| $p_F$, Probability of a adult female reproducing in a year                | 0.5                  | Could be as high at 1, based on field observations                 |
| $F$, Number of surviving eggs produced by an adult female                 | 100                  | From observations of captive frogs                                 |
| $\omega$, Probability of juvenile successfully recruiting to an adult     | Varies yearly        | Explored different values of this parameter                        |

>>>>

---
bibliography: translocation.bib
csl: pnas.csl
format: pdf
# format: docx
# format :
#  latex:
#    cite-method: natbib
editor: visual
editor_options:
  markdown:
    wrap: sentence
---

<!--# Drop "evolutionary rescue" from title? -->

@@

One consequence of human alteration of the biosphere is an increase in emerging infectious diseases [@jones2008; @fisher2012].
Such diseases pose a severe threat to wildlife populations [@daszak2000], and have caused dramatic declines and extinctions in a wide range of taxa, including echinoderms, mammals, birds, and amphibians [@hewson2014; @samuel2015; @scheele2019; @cunningham2021].
Amid this global loss of biodiversity, some populations decimated by disease have subsequently recovered to varying extents [@scheele2017; @voyles2018; @knapp2016].
Although we know relatively little about the mechanisms underlying recovery [@brannelly2021; @russell2020], host evolution is one possibility [@carlson2014; @searle2020].
This process is known from a diversity of wildlife populations [@savage2016; @epstein2016; @gignoux-wolfsohn2021; @holland2022], and often results in increased host resistance and/or tolerance [see @medzhitov2012 for definitions. However, linking these evolutionary responses to population recoveries is difficult and few clear examples exist. In addition to promoting recovery of disease-impacted populations, the resistant/tolerant individuals in such populations could potentially be used to reestablish populations that were extirpated following disease emergence [@mendelson2019; @brannelly2021].
To date, the use of resistant/tolerant individuals for species recovery is rare [but see @joseph2018], and studies that evaluate its potential utility are critically needed.
(In this paper, we do not distinguish between resistance and tolerance because although these terms have different meanings and consequences for host-pathogen dynamics, these distinctions have little significance for our primary focus, i.e., species recovery.) <!--# Feels clumsy to include this paranthetically, but I have not yet come up with a better alternative.  -->

@@

Despite the catastrophic impact of Bd on MYL frogs, wherein most Bd-naive populations are extirpated following Bd arrival [@vredenburg2010], some populations have persisted after epizootics [@briggs2010] and are now recovering [@knapp2016].
Frogs in these recovering populations show reduced susceptibility to Bd infection, with infection intensities ("loads") of adults consistently in the low-to-moderate range regardless of frog size [@ellison2019], frog density [@joseph2018; @wilber2022], or time of year [@knapp2011].
This reduced susceptibility is most consistent with natural selection for more resistant/tolerant genotypes and/or acquired immunity [@grogan2018; @knapp2016], and additional investigation of the specific role of evolution is critically needed.
Exploratory efforts to reintroduce frogs collected from recovering populations to sites from which frogs were previously extirpated following Bd epizootics provide a rare, but preliminary, indication that population establishment is possible despite ongoing Bd infection [@joseph2018].

<!--# I've reworded the next paragraph to provide explicit objectives of each of the sections (plus summary of results). -->

In the current study, we had three primary objectives.
First, to determine whether the reintroduction of resistant/tolerant MYL frogs obtained from populations recovering from Bd-caused declines [@knapp2016] allows the successful reestablishment of extirpated MYL frog populations despite ongoing disease, we conducted a 15-year landscape-scale frog reintroduction effort.
Second, to extend our inferences of population recovery well beyond the temporal extent of our multi-year survey data, we developed a model to estimate the probability of persistence for the reintroduced populations over a multi-decadal period.
Third, we used exome capture methods to provide insights into the evolutionary response by MYL frogs to Bd infection, and thereby provide a mechanism by which frogs reintroduced into sites from which they were previously extirpated by Bd could recover (instead of being driving to extirpation again).
Our results indicate the capacity of reintroduced populations to become established and eventually recover despite ongoing disease, that the recovering populations are likely to persist over a 50-year period, and that frogs have undergone substantial evolutionary change associated with Bd exposure, changes that might provide frogs with increased resistance/tolerance to disease.

<!--# Previous final paragraph of Introduction (revised post-review to reflect reordered sections): In the current study, we combine results from a 15-year landscape-scale frog reintroduction effort, a detailed population model, and genomic analyses to provide an unprecedented example of how reintroduction of resistant/tolerant frogs obtained from recovering populations [@knapp2016] allows the successful reestablishment of extirpated MYL frog populations despite ongoing disease. Specifically, the reintroduction of resistant/tolerant frogs to 24 sites and results from long-term capture-mark-recapture (CMR) surveys indicate the capacity of reintroduced populations to become established and eventually recover. We used these data to parameterize a population model that extends our inferences well beyond the temporal extent of the CMR data, and shows that the recovering populations are likely to persist over a 50-year period. Finally, genomic results obtained using exome capture methods highlight evolutionary rescue as potentially key mechanism by which MYL frogs are able to recover in the presence of ongoing Bd infection. These results show that frogs have undergone substantial evolutionary change associated with Bd exposure, changes that might provide frogs with increased resistance/tolerance to disease. Collectively, these results provide one of the first examples of how evolutionary rescue might facilitate the successful reintroduction of resistant/tolerant animals and allow the large-scale recovery of disease-impacted species.-->

@@

<!--# I moved this paragraph to the top of results to provide an early focus on outcomes that are more obviously related to population recovery than survival. -->

@@

Reintroduction of resistant/tolerant MYL frogs was remarkably successful in reestablishing viable populations in the presence of Bd.
Of the 12 translocated populations, approximately 80% showed evidence of both successful reproduction and recruitment of new adults.
Bd loads measured before and after translocation were similar and generally remained well below the level at which frog mortality occurs.
Annual survival for 12 of the 24 translocated cohorts exceeded 50%, and \> 70% of translocated cohorts had survival above this 50% level when the earliest translocations are excluded (i.e., translocations conducted when methods were still being refined).
The fact that the relatively low Bd loads and correspondingly high frog survival in the presence of Bd was maintained between the natal/donor sites and the recipient sites indicates that these characteristics of naturally-recovering populations were not solely an effect of the natal/donor site or site \* frog interactions, but instead were due at least in part to resistance/tolerance inherent in the frogs <!--# Add to this to deal with Reviewer 2-Comment 4.-->.
In addition, the relatively high survival of translocated frogs was maintained in their progeny, as expected if resistance/tolerance has a genetic basis.
Results from the population viability model were also encouraging.
In particular, translocated populations with \> 50% survival in the first year post-translocation were predicted to have a low probability of extinction over 50 years (probability of extinction = 0 - 0.32).
In light of the generally low success rate of amphibian reintroduction efforts [@dodd2005], our success in reestablishing MYL frog populations via translocation of resistant/tolerant individuals is striking, and even more so given that this success occurred despite ongoing infection of frogs with Bd.
<!--# Mention that 1-year survival can be used as an accurate proxy of long-term persistence.  -->

Despite the resistance/tolerance of adult MYL frogs in recovering populations, individual and population-level impacts of Bd are still evident.
In an earlier study of 2 of our 12 translocated populations [@joseph2018], Bd infection and load had detectable effects on the survival of adults and may have influenced population establishment (sites referred to as "Alpine" and "Subalpine" in @joseph2018 are identified as "70550" and "70505" in the current study).
Applying similar analyses to all 12 of our translocated populations would likely provide a broader perspective of the ongoing effect of Bd.
In addition to these important but relatively subtle effects of Bd on adults, the impacts on younger life stages are more apparent.
MYL frogs immediately following metamorphosis ("metamorphs") are highly susceptible to Bd infection [@ellison2019] and as a result experience high mortality.
This high susceptibility of metamorphs is documented in numerous species of anurans, and may result from the poorly developed immune system characteristic of this life stage [@humphries2022].
In naturally recovering and translocated MYL frog populations, we suggest that the high mortality of metamorphs is an important limitation on subsequent recruitment of new adults.
Therefore, although adult MYL frogs appear relatively resistant/tolerant, Bd infection continues to have important limiting effects on recovering populations [see also @hollanders2022].

@@

We conducted `r transno_min`--`r transno_max` translocations per site (@fig-yosemap, @fig-translocation-survival) and each translocated cohort included `r transfrogs_min` to `r transfrogs_max` frogs (median = `r transfrogs_median`).
In preparation for each translocation, adult frogs were collected from the donor population and measured, weighed, swabbed, and PIT tagged.
Frogs were transported to the recipient site either on foot or via helicopter.
Following release, translocated populations were visited approximately once per month during the summer active season (generally July-August; median number of visits per summer = `r round(surveys_yr_median)`; range = `r surveys_yr_min`--`r surveys_yr_max`) and assessed using diurnal CMR surveys and VES.
CMR surveys allowed estimation of adult survival, adult population size, and recruitment of new adults, and VES provided estimates of tadpole and juvenile abundance.
During 2006-2012, CMR surveys were conducted on a single day (primary period) per site visit, during which all habitats were searched repeatedly for adult frogs.
Frogs were captured using handheld nets, identified via their PIT tag (or tagged if they were untagged), measured, weighed, swabbed, and released at the capture location.
During 2013-2022, we generally used a robust design in which all habitats were searched during several consecutive days (median number of secondary periods per primary period = `r round(robust_year_median)`; range = `r robust_year_min`--`r robust_year_max`), and frogs were processed as described above.
However, when the number of frogs detected on the first survey day was zero or near zero, we typically conducted only a single-day CMR survey.
When using a robust design, within a primary period, frogs that were captured during more than one secondary period were measured, weighed, and swabbed during the first capture, and during subsequent captures were only identified and released.

@@

For each of the translocated populations, we used the presence of tadpoles and/or juveniles from VES and counts of new recruits (i.e, untagged adults) in CMR surveys to provide two measures of successful reproduction.
To calculate the proportion of years in which tadpoles/juveniles were present at a site, we excluded surveys conducted in the year of the initial translocation to that site.
This exclusion accounted for the fact that all translocations were conducted after the breeding period and reproduction would therefore not have occurred until the following year.
Similarly, to calculate the proportion of years in which new recruits were present at a site, we excluded surveys conducted during the 3 years following the initial translocation.
This accounted for the multi-year tadpole and juvenile stages in MYL frogs ([@tbl-param_values]).

@@

@@

![Evidence for selection on individual variants in recovering MYL frog populations at the landscape scale. For each of the 9 naive and recovering MYL frog populations (indicated by numbered points), adjacent pie charts show allele frequencies for the 11 outlier SNPs from 7 distinct genes: (A) LOC108802036, (B) TCF19, VARS, MAP1S, TYRP1, and SRFBP1, and (C) RIN3. Charts are superimposed on a map of the Sierra Nevada study area, with Yosemite, Kings Canyon, and Sequoia National Parks (from north to south) shown in dark gray, and the range boundary for MYL frogs shown in light gray. The inset map locates the study area and range boundary in California. For cross-referencing of numbered points to populations/sites mentioned in the **Frog Population Recovery** section (above), site identifiers for populations 1 and 4 are 72996 and 70567, respectively.](figures/allele_maps.png){#fig-allelefrequencies width="450"}

@@

@@

@@

![Map showing the locations of translocated and donor MYL frog populations in Yosemite National Park (park boundary indicated by gray polygon). Symbol shapes indicate the donor population used for each translocation site, and 5-digit numbers identify each donor and translocation site. To obscure the exact locations of populations, random noise was added to all point coordinates. Inset map shows the location of Yosemite within California. In both maps, elevation is indicated by the colored hillshade layer (dark green = lowest elevation, white = highest elevation).](figures/map_translocation_points.png){#fig-yosemap width="400"}

@@

![Evidence that frogs translocated from recovering populations can maintain the benefits of resistance/tolerance in non-natal habitats <!--# Consider edits to this to respond to Reviewer 2 - Comment 4. In Discussion?-->. For frogs translocated to each of the 12 recipient sites, Bd loads for the period immediately prior to translocation versus during the 1-year period after translocation. Bd loads are expressed as the number of ITS1 copies per skin swab, as estimated by qPCR of the Bd ITS1 region. Box plots show medians, first and third quartiles, largest and smallest values within 1.5x interquartile range, and values outside the 1.5x interquartile range. Loads indicative of severe disease are \> 5.8 ITS copies (on a log~10~ scale). Samples sizes are provided just above the x-axis.](figures/bdload_beforeafter.png){#fig-bdload-beforeafter width="700"}

@@

| **Parameter**                                                             | **Value**            | **Source**                                                         |
|:--------------------------------------------------------------------------|:---------------------|:-------------------------------------------------------------------|
| $\sigma_{L_1}$, Yearly survival probability of year 1 tadpoles            | 0.7                  | Estimated from field data, observations, natural history knowledge |
| $\sigma_{L_2}$, Yearly survival probability of year 2 tadpoles            | 0.7                  | Estimated from field data, observations, natural history knowledge |
| $\sigma_{L_3}$, Yearly survival probability of year 3 tadpoles            | 0.7                  | Estimated from field data, observations, natural history knowledge |
| $\sigma_{J_1}$, Yearly survival probability of year 1 juveniles           | 0.25                 | Estimated from field data, observations, natural history knowledge |
| $\sigma_{J_2}$, Yearly survival probability of year 2 juveniles           | 0.5                  | Estimated from field data, observations, natural history knowledge |
| $\sigma_{A_R}$, Yearly survival probability of naturally recruited adults | Varies by population | Estimated from mark-recapture studies                              |
| $\sigma_{A_T}$, Yearly survival probability of translocated adults        | Varies by population | Estimated from mark-recapture studies                              |
| $p_{L_1}$, Probability of a year 1 tadpoles remaining as a tadpoles       | 1                    | Estimated from field data, observations, natural history knowledge |
| $p_{L_2}$, Probability of a year 2 tadpoles remaining as a tadpoles       | 0.25                 | Estimated from field data, observations, natural history knowledge |
| $p_{J_1}$, Probability of a year 1 juvenile remaining as a juvenile       | 0.25                 | Estimated from field data, observations, natural history knowledge |
| $p_F$, Probability of a adult female reproducing in a year                | 0.5                  | Could be as high at 1, based on field observations                 |
| $F$, Number of surviving eggs produced by an adult female                 | 100                  | From observations of captive frogs                                 |
| $\omega$, Probability of juvenile successfully recruiting to an adult     | Varies yearly        | Explored different values of this parameter                        |
